---
title: "Random Sampling"
author: "Charles Hefer"
date: "`r Sys.Date()`"
output: 
  bookdown::html_document2:
    toc: true
    toc_depth: 4
    df_print: paged
    code_folding: hide
    fig_caption: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if(!require("tidyverse")) {
  install.packages("tidyverse")
}
library(tidyverse)

if (!require("readxl")) {
  install.packages("readxl")
}
library("readxl")

if (!require("writexl")) {
  install.packages("writexl")
}
library("writexl")


```

# A Random Sampler

Read the input file.

```{r}
samples <- readxl::read_xlsx("GFC project and batch info milk species comparison.xlsx",
                           sheet="Sample List for all samples",
                           range="B4:G616") %>%
  mutate("Batch" = NA)
samples
```

The aim is to generate a balanced randomised sample sheet for GFC analysis.

How many species?

```{r}
samples %>%
  group_by(Species) %>%
  summarise(n=n())
```

# Duplicate Injections

Randomly pick 36 samples, 9 from each species that will serve as duplicated injection samples.


```{r}
duplicate_injections <- data.frame()

for (species in c("Cow", "Goat", "Deer", "Sheep")) {
  #print(species)
  selected_samples <- samples %>%
    dplyr::filter(Species == Species) %>%
    dplyr::sample_n(9)
  duplicate_injections <- rbind(duplicate_injections, selected_samples)
}

duplicate_injections

# I want these duplicate injection to be distributed evenly across the 18 batches
duplicate_injections_batched <- data.frame()

for (i in seq(1:18)) {
  batched_injections_selected <- duplicate_injections %>%
    dplyr::sample_n(2) %>%
    mutate(Batch = paste0("B", i)) %>%
    mutate(`Sample Code` = paste0(`Sample Code`, "-dupl"))
  duplicate_injections_batched <- rbind(duplicate_injections_batched,
                                        batched_injections_selected)
}

duplicate_injections_batched
```

Randomly assign the samples to each of the 18 batches

```{r}
random_sample_batches <- data.frame()

for (i in seq(1,18)) {
  batched_samples <- samples %>%
    dplyr::sample_n(36) %>%
    mutate(Batch=paste0("B", i))
  random_sample_batches <- rbind(random_sample_batches, batched_samples)
}

random_sample_batches

```

Merge the injections with the samples

```{r}
batched_samples <- rbind(random_sample_batches,
                       duplicate_injections_batched) %>%
  arrange(Batch)


batched_samples
```

Now, randomly assign the order of the 38 samples in the batch.

```{r}
final_samples <- data.frame()

for (batch in unique(batched_samples$Batch)) {
  print(batch)
  random_samples <- batched_samples %>%
    dplyr::filter(Batch == batch) %>%
    dplyr::sample_n(36)
  
  final_samples <- rbind(final_samples, random_samples)
}

final_samples
```
```{r}
final_samples %>%
  ggplot(aes(x=Species, fill=Batch)) +
  geom_histogram(stat="count")
```

```{r}
final_samples %>%
  ggplot(aes(x=Batch, fill=Species)) + 
  geom_histogram(stat="count")
```


```{r}
write.csv(final_samples,
          file="Randomised_samples.csv")
```

