---
title: "Quality Control"
editor: source
---

```{r, include=FALSE, warning=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)

#source some global libraries, from globals.R
source("globals.R")
theme_set(theme_bw())

base::load("01_ExperimentSetup.Rdata")

```

# Quality Control

Make use of several plots to visualize the pattern for missing values, the number of proteins identified, and the number of proteins that are common or unique across the samples.

PCA is the main tool to evaluate the relationship between samples and the QCMIX control samples.

## Missing value patterns

Missing values is a known problem within proteomics data. Too many missing values causes doubt about the real abundance values for peptides. Below is a plot of the dataset with no missing values, as well as the filtered dataset where the filter requirement is a protein that is measured in at least 75% of the samples, for a specific group.

```{r}
#| label: plot_missval
#| layout-ncol: 2
#| fig-cap: Missing value pattern for the dataset
#| fig-subcap: 
#|   - Unfiltered data
#|   - Filtered Dataset
#| warning: False
#| message: False
#| cache: True

plot_missval(SE)

#as an illustration, filter based on the condition
filt_SE <- filter_proteins(SE, type="condition", thr=0.25)
plot_missval(filt_SE)
```



### Number of Proteins per sample

Plot the number of proteins detected in each of the samples

```{r}
#| label: plot_proteins_per_sample
#| layout-ncol: 2
#| fig-cap: Number of proteins per sample.
#| fig-subcap: 
#|   - Unfiltered data
#|   - Filtered Dataset
#| warning: False
#| message: False


plot_numbers(SE)
plot_numbers(filt_SE)
```

### Abundance distribution

Inspect the distribution of abundance levels between samples.

```{r warning=FALSE}
#| label: plot_abundance
#| layout-ncol: 2
#| fig-cap: Protein abundance distribution
#| fig-subcap: 
#|   - Unfiltered data
#|   - Filtered Dataset
#| warning: False
#| message: False


get_df_wide(SE) %>%
  #assign the ID to the rownames
  magrittr::set_rownames(.$ID) %>%
  #filter the name and ID columns
  dplyr::select(-c(ID, name)) %>%
  t() %>%
  data.frame() %>%
  mutate(sample = rownames(.)) %>%
  mutate(group = METADATA$condition) %>%
  pivot_longer(names_to="protein", values_to="abundance", -c(sample, group)) %>%
  ggplot(aes(x=sample, y=abundance, fill=group)) + 
    geom_boxplot() + 
    theme_bw() +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1)) + 
    xlab("") + 
    ylab("log2(abundance)")


get_df_wide(filt_SE) %>%
  #assign the ID to the rownames
  magrittr::set_rownames(.$ID) %>%
  #filter the name and ID columns
  dplyr::select(-c(ID, name)) %>%
  t() %>%
  data.frame() %>%
  mutate(sample = rownames(.)) %>%
  mutate(group = METADATA$condition) %>%
  pivot_longer(names_to="protein", values_to="abundance", -c(sample, group)) %>%
  ggplot(aes(x=sample, y=abundance, fill=group)) + 
    geom_boxplot() + 
    theme_bw() +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1)) + 
    xlab("") + 
    ylab("log2(abundance)")
```



## PCA QCMix samples

Perform a PCA analysis for the experiment, with a focus in the reproducibility of the QC_Mix samples. This is being done on the filtered dataset.



```{r}
#| label: get_df


lfq <- get_df_wide(filt_SE) %>%
  #assign the ID to the rownames
  magrittr::set_rownames(.$ID) %>%
  #filter the name and ID columns 
  dplyr::select(-c(name, ID)) %>%
  t()

#filter out proteins that were in less than 0.75 of the samples
#One-hit wonders....
#t_lfq <- t(lfq)
#lfq_filt <- t(t_lfq[rowSums(ifelse(is.na(t_lfq),1,0)) < round(ncol(t_lfq)*0.5),])

#validate that the lfq rows match the expr cols
rownames(lfq) == METADATA$name
```

```{r}
#| label: perform_pca
#| cache: True

#Perform the PCA
pca_lfq <- pca(lfq,
              ncomp=3,
              center=T,
              scale=F)
```

From the original `r dim(lfq)[2]` proteins, filtered out based in their missingness in the experiment. These proteins were removed prior to performing the PCA.


```{r}
#| label: pca_plot_generation
#| include: FALSE
#| echo: FALSE

plot1 <- plotIndiv(pca_lfq,
          comp=c(1,2),
          group=METADATA$condition,
          ellipse=F,
          title="QC PCA (comp 1,2)",
          legend=F)

plot2 <- plotIndiv(pca_lfq,
          comp=c(2,3),
          group=METADATA$condition,
          ellipse=F,
          title="QC PCA (comp 2,3)",
          legend=F)
```

```{r fig.cap="Principle Component Analysis on the dataset, including the QC_Mix samples. Figure (a) represents the first 2 components, and figure (b) components 2 and 3."}
plot_grid(plot1$graph,
          plot2$graph,
          ncol=2,
          nrow=1,
          labels="auto",
          rel_widths = c(1,1))
```

The QCMix samples cluster as expected. There is quite a bit of biological variation, with especially sample `None_None_1` being an outlier in the dataset.

These samples are being used for the `Sample LFQ` analysis that follows next.

```{r}
#save the QC-ed dataset, if there were any changes
METADATA <- metadata
save(METADATA, SE, MAXLFQ, file="02_QualityControl.Rdata")
```
