---
title: "Quality Control"
editor: source
---

```{r, include=FALSE, warning=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE)

#source some global libraries, from globals.R
source("globals.R")
theme_set(theme_bw())

base::load("01_ExperimentSetup.RDS")

```

# Quality Control

Make use of several plots to visualize the pattern for missing values, the number of proteins identified, and the number of proteins that are common or unique across the samples.

PCA is the main tool to evaluate the relationship between samples and the QCMIX control samples.

## Missing value patterns

Missing values is a known problem within proteomics data. Too many missing values causes doubt about the real abundance values for peptides. Plot the missing value pattern for the `r dim(SE)[1]` detected proteins in @fig-plot_missval.

```{r}
#| label: fig-plot_missval
#| layout-ncol: 1
#| fig-cap: Missing value pattern for the dataset
#| warning: False
#| message: False
#| cache: True

plot_missval(SE)

#as an illustration, filter based on the condition
#filt_SE <- filter_proteins(SE, type="condition", thr=0.25)
#plot_missval(filt_SE)
```


:::{.callout-note}
Typically we do not expect the biological samples to cluster together, unless there is a specific condition where a protein will be absent in that condition. However, this will be a small subset of the proteins identified.
:::


### Number of Proteins per sample

The total number of proteins detected is plotted as a histogram in @fig-plot_proteins_per_samnple.

```{r}
#| label: fig-plot_proteins_per_sample
#| layout-ncol: 1
#| fig-cap: Number of proteins per sample.
#| warning: False
#| message: False


plot_numbers(SE)
#plot_numbers(filt_SE)
```

### Abundance distribution

Inspect the distribution of abundance levels between samples in @fig-plot_abundance.

```{r warning=FALSE}
#| label: fig-plot_abundance
#| layout-ncol: 1
#| fig-cap: Protein abundance distribution (log2)
#| warning: False
#| message: False


get_df_wide(SE) %>%
  #assign the ID to the rownames
  magrittr::set_rownames(.$ID) %>%
  #filter the name and ID columns
  dplyr::select(-c(ID, name)) %>%
  t() %>%
  data.frame() %>%
  mutate(sample = rownames(.)) %>%
  mutate(group = METADATA$condition) %>%
  pivot_longer(names_to="protein", values_to="abundance", -c(sample, group)) %>%
  ggplot(aes(x=sample, y=abundance, fill=group)) + 
    geom_boxplot() + 
    theme_bw() +
    theme(axis.text.x = element_text(angle=90, vjust=0.5, hjust=1)) + 
    xlab("") + 
    ylab("log2(abundance)")

```



## Principle Component Analysis (PCA)

Perform a PCA analysis for the experiment, with a focus in the reproducibility of the QC_Mix samples.


```{r}
#| label: get_df
#| include: FALSE
#| echo: FALSE

lfq <- get_df_wide(SE) %>%
  #assign the ID to the rownames
  magrittr::set_rownames(.$ID) %>%
  #filter the name and ID columns 
  dplyr::select(-c(name, ID)) %>%
  t()

#validate that the lfq rows match the expr cols
rownames(lfq) == METADATA$name
```

```{r}
#| label: perform_pca
#| cache: TRUE

#Perform the PCA
pca_lfq <- pca(lfq,
              ncomp=3,
              center=T,
              scale=F)
```

```{r}
#| label: pca_plot_generation
#| include: FALSE
#| echo: FALSE
#| warning: FALSE

plot1 <- plotIndiv(pca_lfq,
          comp=c(1,2),
          group=METADATA$condition,
          ellipse=F,
          title="QC PCA (comp 1,2)",
          legend=F)

plot2 <- plotIndiv(pca_lfq,
          comp=c(2,3),
          group=METADATA$condition,
          ellipse=F,
          title="QC PCA (comp 2,3)",
          legend=F)
```

```{r fig.cap="Principle Component Analysis on the dataset, including the QC_Mix samples. Figure (a) represents the first 2 components, and figure (b) components 2 and 3."}
plot_grid(plot1$graph,
          plot2$graph,
          ncol=2,
          nrow=1,
          labels="auto",
          rel_widths = c(1,1))
```

The `QCMix` samples cluster as expected in the center position of the plot. The `W2_LB` samples cluster nicely together, the same for `D0_H20` and `W2_C`. `D0_C` and `W2_H20` is not separated in the first or second component, but on the third. `D0_LB` shows a lot of variation.

Overall, PC1 explains 16% of the variance, PC2 10%, and PC3 9%. The samples are very similar.

```{r}
#| label: save_qcmix_samples
#save the QC-ed dataset, if there were any changes
METADATA <- metadata
save(METADATA, SE, LFQ, LFQ.ORIG, file="02_QualityControl.Rdata")
```
